Rationality, Regret, Evolution \& the Environment

### 2015-11-05 paper/paper.tex
Paolo tweaks to Sec 3.3 and beyond

@@

Consider the following problem. I have, on the desk in front of me, a bag containing either red or black marbles. There is no further information about the distribution of the two colors. What is then the probability that I will draw a red marble? As in Walley91-96, we are not talking about physical probability, but rather about epistemic probability, i.e., the subjective probability that an agent attributes to an event.
In the literature there are two traditional ways of answering this kind of questions. According to standard Bayesianism, one should always be able to determine a precise betting rate (that immediately translates into a precise probability distribution), which is fair, in the sense that the DM is indifferent between betting on or against red at the fair rate, and would bet on red at any rate that is more favourable than the fair rate and against red at any rate that is less favourable. In this case, a strict Bayesian, in the absence of any further information, would probably rely on the principle of insufficient reason and answer that the DM should have a uniform distribution that considers all the outcomes equipossible. \\
The second answer involves imprecise probabilities and is related to the literature about unmeasurable uncertainty. Many authors argued in favor of this kind of approach [e.g., Walley91-96, Levi74, GardenforsANDSahlin82]. Since it is consistent with the information available that all the marbles are red, or that no marble is red, the resulting model should take it into account and specify a lower probability of red $\underline{P}(R)$ and an upper probability of red $\overline{P}(R)$ that define an interval of possible probabilistic beliefs [$\underline{P}(R), \overline{P}(R)$]. Given the information available in this case we would have $\underline{P}(R)=0$ and $\overline{P}(R)=1$. This approach appears particularly relevant and appropriate in a game theoretical context. Indeed, in a recent paper [Battigalliet al.15] Battigalli et al. write

@@

In decision theory, the imprecise probability model is in line with most of the representation results about uncertainty [GilboaSchmeidler89,KMM06], and seems justified by empirical observations too.

@@

A huge body of experimental literature strongly suggests that agents are generally uncertainty
averse [REF]. In line with empirical data,
we assume that players in the population are uncertainty averse and conform to maxmin expected
utility. The resulting behavior is then produced by maximinimizing the subjective utility
given by the player type over the set of probabilities given by the belief state, or epistemic type, of the agent.

In full generality, we think of an epistemic type as a general disposition to form beliefs about the co-player’s behavior.    Players in the population obviously face uncertainty about the composition of the population that they are part of, and consequently about the type of co-player that they are randomly paired with at each round and about the co-player's actions. We represent this uncertainty in two different ways that correspond to the two approaches discussed above: a Bayesian uniform distribution and an imprecise probability set. This amounts to having two different epistemic types, $ \overline{\mu}$ and $ \Delta(A_{-i}) \rbrace $, where $\overline{\mu}$ stands for the uniform distribution and $\Delta(A_{-i})$ is the probability set, defined as follows:

Hence, we consider two possible epistemic states in the face of uncertainty. The epistemic type $\Delta(A_{-i})$ represents a situation of non-Bayesian radical uncertainty, where the players are not able to narrow down the set of probability distributions to a single one and they consider all of them as possible. The other epistemic type $\overline{\mu}$ conceptualizes the uncertainty about the population by using the principle of insufficient reason and ascribes equal probability to all the possible alternatives. As we have seen, both options are reasonable and justifiable ways to deal with situations of uncertainty [REFERENCES]. \\

@@

Fixing the action selection component to maximin expected utility, and allowing two belief and
two preference types, we end up with four basic choice mechanisms summed up in
Table~\ref{tab:CMs}. Different choice mechanisms can give rise to identical action predictions,
either in a restricted class of games, or in general. This is also the case here.

@@

Moreover, our choice mechanism $(\tau^{\pi}, \Delta(A_{-i}))$ turns out to be the objective
maximinimizer, player $(\tau^{reg}, \Delta(A_{-i}))$ is the regret minimizer, player
$(\tau^{\pi}, \overline{\mu})$ is the objective expected utility maximizer, and player
$(\tau^{reg}, \overline{\mu})$ is the expected regret minimizer. In other words, behaviorally,
the choice mechanisms we defined here correspond to the classical choice rules in
Table~\ref{tab:CMs}.

@@

\todo[inline]{Micha rewrite up to here!!! Please check which bits of text/information from the
  following need to be integrated and what can go.}

@@

We are now in the position of deriving behavior directly from the pair $(\tau, e)$. Definitions \ref{def:utility-belief} and \ref{def:maxminSEU} simply formalize the notions of subjective expected utility and maxmin subjective expective utility, respectively.

\noindent Definition \ref{def:expected fitness} gives the expected evolutionary fitness of player $(\tau, e)$ against player $(\tau', e')$ in game $G$.\footnote{As already specified in the previous example, we assume that, whenever the set $\text{maxmin}(\tau(G), e)$ is not a singleton, the player $(\tau, e)$ will mix with equal probability among the actions in $\text{maxmin}(\tau(G), e)$.}

\begin{definition} \label{def:expected fitness}

@@

\subsection{Epistemic types}
\label{sec:epistemic-types}

\todo[inline]{I think we really need a better (and earlier) justification for these epistemic
  types; we could appeal to information available to agents; naturalness of situations in which
  we cannot attach numbers to our beliefs (life on Mars again) \dots}

Consider the following problem. I have, on the desk in front of me, a bag containing either red or black marbles. There is no further information about the distribution of the two colors. What is then the probability that I will draw a red marble? As in Walley91-96, we are not talking about physical probability, but rather about epistemic probability, i.e., the subjective probability that an agent attributes to an event.
In the literature there are two traditional ways of answering this kind of questions. According to standard Bayesianism, one should always be able to determine a precise betting rate (that immediately translates into a precise probability distribution), which is fair, in the sense that the DM is indifferent between betting on or against red at the fair rate, and would bet on red at any rate that is more favourable than the fair rate and against red at any rate that is less favourable. In this case, a strict Bayesian, in the absence of any further information, would probably rely on the principle of insufficient reason and answer that the DM should have a uniform distribution that considers all the outcomes equipossible. \\
The second answer involves imprecise probabilities and is related to the literature about uncertainty, or ambiguity. Many authors argued in favor of this kind of approach [e.g., Walley91-96, Levi74, GardenforsANDSahlin82]. Since it is consistent with the information available that all the marbles are red, or that no marble is red, the resulting model should take it into account and specify a lower probability of red $\underline{P}(R)$ and an upper probability of red $\overline{P}(R)$ that define an interval of possible probabilistic beliefs [$\underline{P}(R), \overline{P}(R)$]. Given the information available in this case we would have $\underline{P}(R)=0$ and $\overline{P}(R)=1$. This approach appears particularly relevant and appropriate in a game theoretical context. Indeed, in a recent paper [Battigalliet al.15] Battigalli et al. write

In decision theory, the imprecise probability model is in line with most of the representation results about uncertainty [GilboaSchmeidler89,KMM06], and seems justified by empirical observations too (see below). Walley91-96 is a seminal work in the area, that also introduces learning procedures to get useful inferences and narrow down the probability set [$\underline{P}(R), \overline{P}(R)$] after observing data. Since we want to study the evolution of meta-types, the present work does not address the probem of drawing inferences from observations, and for the moment we do not endow the players with any learning procedure.\\
In line with traditional evolutionary game theory, we model a population of rather unsophisticated and myopic agents. %that possibly differ from each other with respect to two components: the player type and the epistemic type. The players 
The players encode both a psychological subjective representation of the interaction (player type) and a conceptualization of uncertainty (epistemic type), but they are too simple to have any learning or belief updating procedure. From this point of view, we can take the learning process away from natural selection and focus on player types and epistemic types as the target of evolutionary pressure.

In full generality, we think of an epistemic type as a general disposition to form beliefs about the co-player’s behavior.    Players in the population obviously face uncertainty about the composition of the population that they are part of, and consequently about the type of co-player that they are randomly paired with at each round and about the co-player's actions. We represent this uncertainty by the use of both the approaches presented above: a uniform distribution and a probability set. This amounts to having two different epistemic types $e$ in the population, $e \in \lbrace \overline{\mu}, \Delta(A_{-i}) \rbrace $, where $\overline{\mu}$ stands for the uniform distribution and $\Delta(A_{-i})$ is the probability set, defined as follows:

\item $\overline{\mu} \in \lbrace \mu \in \mathbb{R}^{A_{-i}}: \sum_{a_{-i} \in A_{-i}} \mu(a_{-i}) = 1 \rbrace$ is the probability measure over the co-player's actions such that: $\forall a_{-i} \in A_{-i}$, $\overline{\mu}(a_{-i})= \frac{1}{|A_{-i}|}$.

Hence, we consider two possible epistemic states in the face of uncertainty. The epistemic type $\Delta(A_{-i})$ represents a situation of non-Bayesian radical uncertainty, where the players are not able to narrow down the set of probability distributions to a single one and they consider all of them as possible. The other epistemic type $\overline{\mu}$ conceptualizes the uncertainty about the population by using the principle of insufficient reason and ascribes equal probability to all the possible alternatives. As we have seen, both options are reasonable and justifiable ways to deal with situations of uncertainty [REFERENCES]. \\

\subsection{Uncertainty aversion and maxmin rule}

\todo[inline]{this paragraph seems to presuppose familiarity with maxi-minimization (can we
  write this with a hyphen, any chance?); we should rewrite this or introduce the notion immediately;}

An important part of the latest literature, both experimental and theoretical, in decision theory deals with the issue of uncertainty, also called ambiguity. Technically, there is uncertainty about an event when no precise objective probability distribution about the possible outcomes is given. A prominent example is the above mentioned Ellsberg's paradox. A bag contains 90 marbles, 30 are red and the remaining 60 are either black or yellow. A marble is to be drawn and DM can win 100€ if she guesses the color of the drawn marble. The DM is offered two different bets. In the first one she can choose to bet either on red (R) or on yellow (Y), while in the second she can bet either on red or black (RB) or on yellow or black (YB), as shown in Table \ref{Ellsberg}.

The pattern consistently observed in experimets is: $f_{{\color{red}R}} \succ f_{{\color{yellow}Y}}$ and $f_{{\color{red}R}B} \prec f_{{\color{yellow}Y}B}$. This is clearly incompatible with any precise probabilistic belief about the proportion of black and yellow marbles. This kind of behavior has been
famously axiomatized by Gilboa and Schmeidler in a paper from 1989. The choice pattern is normally explained in terms of aversion to uncertainty, and the standard representation is given by means of probability sets together with maxmin rule. We can think of the DM in Ellsberg's example as maximinimizing expected utility over the probability set $ [\underline{P}(R)=\frac{1}{3},\underline{P}(Y)=0,\underline{P}(B)=0,\overline{P}(R)=\frac{1}{3},\overline{P}(Y)=\frac{2}{3},\overline{P}(B)=\frac{2}{3}]  $. This amounts to calculating for any probability distribution contained in the probability set the expected utility of any available action, and then to choosing the action that guarantees the highest minimal expected utility. Whenever the set of priors is a singleton, maximinization of
expected utility coincides with maximization of expected utility.\\
%Given a player type and an epistemic type, there are different ways to produce a behavior. 
A huge body of experimental literature strongly suggests that agents are generally uncertainty
(or ambiguity) averse [REF]. In line with empirical data,
we assume that players in the population are uncertainty averse and conform to maxmin expected
utility. The resulting behavior is then produced by maximinimizing the subjective utility
given by the player type over the set of priors given by the epistemic type.

  \todo[inline]{I would like to suggest to move Figure 2 next to Figure 1; it is already useful
  for example 1 and could be referred to there already; I would suggest to use subfigures to
  put both next to each other}

\begin{figure}
\begin{center}%
\begin{tabular}{|c|c|c|}
\hline 
 & I & II\tabularnewline
\hline 
\hline 
I & 0;0 & -2;-1\tabularnewline
\hline 
II & -1;-2 & 0;0\tabularnewline
\hline 
\end{tabular}\end{center}

\protect\caption{Regret representation $\tau^{reg}(G^C)$}
\label{coordgame1reg}
\end{figure}

We are now in the position of deriving behavior directly from the pair $(\tau, e)$. Definitions \ref{def:utility-belief} and \ref{def:maxminSEU} simply formalize the notions of subjective expected utility and maxmin subjective expective utility, respectively.

\noindent Definition \ref{def:expected fitness} gives the expected evolutionary fitness of player $(\tau, e)$ against player $(\tau', e')$ in game $G$.\footnote{As already specified in the previous example, we assume that, whenever the set $\text{maxmin}(\tau(G), e)$ is not a singleton, the player $(\tau, e)$ will mix with equal probability among the actions in $\text{maxmin}(\tau(G), e)$.}

\begin{definition} \label{def:expected fitness}

Many other alternatives different from maxmin expected utility could produce behavior: choice rules like minmax, maxmax, and minmin expected utility would similarly generate behavior from a player type and an epistemic type. Here we postulate that players are maximinimizers for two main reasons. First, as already mentioned, among other possible choice rules maxmin is not only intuitively appealing but also corroborated from an empirical point of view. Secondly, our main interest is the evolution of different patterns of subjective conceptualizations of games (player types), and we want to show that the veridical representation (objective type) is not necessarily the fittest and may not be evolutionarily stable. Assuming that all players use the maxmin rule independent of their player type allows us to factor the choice rule out of the selection process and to focus on player types.

Before concluding this section, it is worth stressing that player $(\tau^{\pi}, \Delta(A_{-i}))$ turns out to be the objective maximinimizer, player $(\tau^{reg}, \Delta(A_{-i}))$ is the regret minimizer, player $(\tau^{\pi}, \overline{\mu})$ is the objective expected utility maximizer, and player $(\tau^{reg}, \overline{\mu})$ is the expected regret minimizer. Moreover, the following facts hold.

\noindent As we have seen in the previous example, this is not the case in the presence of (radical) uncertainty, in the sense of representation of belief state in terms of a set of probability, as in $ \Delta(A_{-i}) $.

\begin{fact} \label{fact:equivalence2x2}

In the class of $2 \times 2$ symmetric games, players $(\tau^{reg}, \Delta(A_{-i}))$, $(\tau^{\pi}, \overline{\mu})$, and $(\tau^{reg},\overline{\mu})$ are behaviorally equivalent.

\noindent Two remarks. Firstly, Fact \ref{fact:equivalence2x2} does not hold for $n \times n$
symmetric games in general. Secondly, we treat behaviorally equivalent types as distinct,
because we assume that what is selected by biological evolution is not choice behavior as such,
but the mechanism which generates behavior. This matters in particular if we assume that
mutations affect different components of the behavior generating mechanisms (see below).

\fi

>>>>

\subsection{Choice mechanisms take2}

In the previous section we introduced choice rules generating behavior in a game as functions from a game to an action choice. In the general form, given a class of games $\mathcal{G}$, a choice rule is then thought as a function $c: \mathcal{G} \rightarrow A_i $. In this section instead we want to show that a more intelligible and perspicuous way of understanding behavior generating mechanisms may be as functions of a subjective preference and an epistemic state of the agents. Then, given a set of subjective preferences $ \Theta $ [@Paolo: minimally define subjective preferences somewhere above] and a set of epistemic states (or belief types) $E$, a choice mechanism is a function $\kappa: \Theta \times E \rightarrow A_i $. We will see in a moment that from this perspective we gain a more unified and structured insight on the choice rules presented above.

More
concretely, we focus on two \emph{preference types} and two \emph{belief types}. We view a preference type as a function $\tau: \mathcal{G} \rightarrow \Theta$, and a belief type as a general disposition to form beliefs about the co-player’s behavior. The two possible dispositions that we take account of are probabilistic (precise) belief and non-probabilistic (imprecise) belief, as we will explain in the next section.

We are most interested here in studying whether evolutionary pressure could support
non-veridical subjective preferences under certain conditions. We therefore consider two
\emph{preference types}: veridical type $\tau^\pi$, who uses the actual payoff function $\pi$ of the
actual game $G$, and regret type $\tau^{reg}$, who looks at the regret transformation of the
actual game $G$.

The choice mechanism that we use in this paper is \textit{maxmin expected utility} [GS89], for reasons that will be clear and justified in the following two sections.

@@

Consider the following problem. I have, on the desk in front of me, a bag containing either red or black marbles. There is no further information about the distribution of the two colors. What is then the probability that I will draw a red marble? As in Walley91-96, we are not talking about physical probability, but rather about epistemic probability, i.e., the subjective probability that an agent attributes to an event.
In the literature there are two traditional ways of answering this kind of questions. According to standard Bayesianism, one should always be able to determine a precise betting rate (that immediately translates into a precise probability distribution), which is fair, in the sense that the DM is indifferent between betting on or against red at the fair rate, and would bet on red at any rate that is more favourable than the fair rate and against red at any rate that is less favourable. In the case of the bag on my desk, a strict Bayesian, in the absence of any further information, would probably rely on the \textit{principle of insufficient reason} and answer that the DM should have a uniform distribution that considers all the outcomes equipossible. \\
The second answer involves imprecise probabilities and is related to the literature about unmeasurable uncertainty. Many authors argued in favor of this kind of approach [e.g., Walley91-96, Levi74, GardenforsANDSahlin82]. Since it is consistent with the information available that all the marbles are red, or that no marble is red, the resulting model should take it into account and specify a lower probability of red $\underline{P}(R)$ and an upper probability of red $\overline{P}(R)$ that define an interval of possible probabilistic beliefs [$\underline{P}(R), \overline{P}(R)$]. Given the information available in this case we would have $\underline{P}(R)=0$ and $\overline{P}(R)=1$. This approach appears particularly relevant in a game theoretical context. Indeed, in a recent paper [Battigalliet al.15] Battigalli et al. write

@@

In evolutionary game theory, for instance, players in a population obviously face uncertainty about the composition of the population that they are part of, and consequently about the co-player that they are randomly paired with at each round and about the co-player's action. We represent this uncertainty in two different ways that correspond to the two approaches discussed above: a Bayesian uniform distribution and an imprecise probability set. This amounts to having two different belief types, $ \overline{\mu}$ and $ \Delta(A_{-i}) \rbrace $, where $\overline{\mu}$ stands for the uniform distribution and $\Delta(A_{-i})$ is the probability set, defined as follows:

Hence, we consider two possible epistemic states in the face of uncertainty. The belief type $\Delta(A_{-i})$ represents a situation of non-Bayesian radical uncertainty, where the players are not able to narrow down the set of probability distributions to a single one and they consider all of them as possible. The other belief type $\overline{\mu}$ conceptualizes the uncertainty about the population by using the principle of insufficient reason and ascribes equal probability to all the possible alternatives. As we have seen, both options are reasonable and justifiable ways to deal with situations of uncertainty [REFERENCES]. \\

In decision theory, the imprecise probability model is in line with most of the representation results in decision making under uncertainty [GilboaSchmeidler89,KMM06], and seems justified by empirical observations too.

@@

Evidence from experimental literature suggests that agents are mainly uncertainty
averse [REF]. In line with empirical data,
we assume that players in the population are uncertainty averse and conform to maxmin expected
utility. The resulting behavior is then produced by maximinimizing the subjective utility
given by the player (preference, UNIFORM TERMINOLOGY CHECK) type over the set of probabilities given by the belief type of the agent.

@@

Fixing the choice mechanism (action selection component) to maximin expected utility, and allowing two belief and
two preference types, we regain all the choice rules considered in sectionNN from a more general and unified standpoint, as it is summed up in
Table~\ref{tab:CMs}. \\
We are now in the position of formally deriving behavior from a pair $(\theta, e) \in \Theta \times E$.

[[\noindent Definition \ref{def:expected fitness} gives the expected evolutionary fitness of player $(\tau, e)$ against player $(\tau', e')$ in game $G$.\footnote{As already specified in the previous example, we assume that, whenever the set $\text{maxmin}(\tau(G), e)$ is not a singleton, the player $(\tau, e)$ will mix with equal probability among the actions in $\text{maxmin}(\tau(G), e)$.}
[OF COURSE we will postpone the definition of evolutionary fitness, it is only momentarily here!]
\begin{definition} \label{def:expected fitness}

\end{definition}
[OF COURSE we will postpone the definition of evolutionary fitness, it is only momentarily here!]]]

It is then worth stressing here that $\kappa(\tau^{\pi}, \Delta(A_{-i}))$ turns out to be the objective
maximinimizer, $\kappa(\tau^{reg}, \Delta(A_{-i}))$ corresponds to the regret minimizer,
$\kappa(\tau^{\pi}, \overline{\mu})$ is the objective expected utility maximizer (with flat beliefs), and
$\kappa(\tau^{reg}, \overline{\mu})$ is the expected regret minimizer (with flat beliefs).
Moreover, different subjective preferences and belief types can give rise to identical action predictions,
either in a restricted class of games, or in general. This is also the case here.

@@

@@

\todo[inline]{PAOLO rewrite up to here!!! BUT the last 2-3 paragraphs must be rearranged a bit more!}

@@

@@
